Fast Sorting on Modern Computers

### 2005-07-31 docs/tech_report/elementary.tex
spell checking. just noticed the word mergsort, so need to fix it

\section{Improved Shakersort}
Shakersorts incorporates the same improvements over shakersort as improved bubblesort
does over bubblesort, that is, it skips over pockets of sorted keys in both
directions, as well as ending early. The code for improved shakersort is in Figure
\vref{Improved shakersort code}.

>>>>

\section{Improved Shakersort}
Improved shakersort incorporates the same improvements over shakersort as
improved bubblesort does over bubblesort, that is, it skips over pockets of
sorted keys in both directions, as well as ending early. The code for improved
shakersort is in Figure \vref{Improved shakersort code}.

### 2005-07-31 docs/tech_report/heap.tex
spell checking. just noticed the word mergsort, so need to fix it

LaMarca overcomes the problem by inserting a maximal key at the end of a heap if
there's an even number of nodes in the array. This gives a second child to the
last element of the heap. This is an effective technique for building the heap,
as the number of keys is static. When destroying the heap, however, the
maximal key must be added to every second operation. This can be avoided
relatively simply with no extra instructions. If the last key in the heap is
left in place aswell as being sifted down the heap, it will provide a
sentinel for the sift-down operation. The key from the top of the heap is then
put into place at the end. This technique reduces instruction count by
approximately 10\%.

@@

In the graph, the base heapsort's results flat-line for a step longer than the
other heapsorts' results. This is due to being in-place, since twice as many
keys fit in the cache without a corresponding increase in misses. For the same
reason, the base heapsorts have no compulsary misses\footnote{See Section
\ref{TODO}.}, while the out-of-place sorts have one compulsory miss in eight,
when the auxilliary heap is being populated.

@@

Slight differences exist though. The level 2 cache miss results here indicate
that while still sorting in the cache, the number of misses in the in-place
heapsort is less than that of the out-of-place sorts. LaMarca, however, has the
same amount of misses in both cases. Similarly, the results here indicate that
base heapsort experiences a sharp increase in cache misses. Cache heaport also
has a similar increase, though LaMarca's results show that these occur at the
same position in the graph, in other words, when th set size is the same. Our
results, however, show that this occurs earlier in memory-tuned heapsort, due to the
fact that it is an out of place sort.

@@

The most interesting thing is the relation between the predictablility of the
comparisons during the destruction of the 2-, 4- and 8-heaps. Compare the
columns marked `\cc{Destruction, child comparison}' and `\cc{Destruction, child
promotion}' in Figure \ref{base and memory-tuned heapsort-1} (the 2-heap), with
the columns marked `\cc{comparison 1}' to `\cc{child promotion}' in Figure
\ref{base and memory-tuned heapsorts-2} (the 4-heap) and Figures \ref{base and
memory-tuned heapsort-3} and \ref{base and memory-tuned heapsort-4} (the
8-heap).

@@

LaMarca later published heapsort code in \cite{LaMarcaHeap96}. The ACM
Journal of Experimental Algorithms requires that source code be published, and
comparing simulated results from there to the sorts created here might have
interesting results and would explain the discreptancies highlighted in this
chapter.

>>>>

LaMarca overcomes the problem by inserting a maximal key at the end of a heap if
there's an even number of nodes in the array. This gives a second child to the
last element of the heap. This is an effective technique for building the heap,
as the number of keys is static. When destroying the heap, however, the
maximal key must be added to every second operation. This can be avoided
relatively simply with no extra instructions. If the last key in the heap is
left in place as well as being sifted down the heap, it will provide a
sentinel for the sift-down operation. The key from the top of the heap is then
put into place at the end. This technique reduces instruction count by
approximately 10\%.

@@

In the graph, the base heapsort's results flat-line for a step longer than the
other heapsorts' results. This is due to being in-place, since twice as many
keys fit in the cache without a corresponding increase in misses. For the same
reason, the base heapsorts have no compulsory misses\footnote{See Section
\ref{TODO}.}, while the out-of-place sorts have one compulsory miss in eight,
when the auxiliary heap is being populated.

@@

Slight differences exist though. The level 2 cache miss results here indicate
that while still sorting in the cache, the number of misses in the in-place
heapsort is less than that of the out-of-place sorts. LaMarca, however, has the
same amount of misses in both cases. Similarly, the results here indicate that
base heapsort experiences a sharp increase in cache misses. Cache heapsort also
has a similar increase, though LaMarca's results show that these occur at the
same position in the graph, in other words, when the set size is the same. Our
results, however, show that this occurs earlier in memory-tuned heapsort, due to the
fact that it is an out of place sort.

@@

The most interesting thing is the relation between the predictability of the
comparisons during the destruction of the 2-, 4- and 8-heaps. Compare the
columns marked `\cc{Destruction, child comparison}' and `\cc{Destruction, child
promotion}' in Figure \ref{base and memory-tuned heapsort-1} (the 2-heap), with
the columns marked `\cc{comparison 1}' to `\cc{child promotion}' in Figure
\ref{base and memory-tuned heapsorts-2} (the 4-heap) and Figures \ref{base and
memory-tuned heapsort-3} and \ref{base and memory-tuned heapsort-4} (the
8-heap).

@@

LaMarca later published heapsort code in \cite{LaMarcaHeap96}. The ACM
Journal of Experimental Algorithms requires that source code be published, and
comparing simulated results from there to the sorts created here might have
interesting results and would explain the discrepancies highlighted in this
chapter.

### 2005-07-31 docs/tech_report/merge.tex
spell checking. just noticed the word mergsort, so need to fix it

For his base algorithm, LaMarca takes Knuth's algorithm N \cite{Knuth98}.
LaMarca describes this algorithm as functioning as described above; lists of
length one are merged into lists of length two, then four, eight, and so on.
However, algorithm N does not work in this fashion. Instead, it works as a
\n{natural} merge. Rather than each list having a fixed length at a certain
point, lists are merged according to the initial ordering. If it happens that
there exists an ascending list on one side, this is exploited by the algorithm.
While this results in good performance, the merge is not perfectly regular, and
it is difficult or impossible to perform several of the optimizations LaMarca
describes on it. We attempted to contact him in order to discuss this, but
recieved no response. It is likely that the improvements were actually applied
to Knuth's algorithm S.

@@

\sixplot{algorithm_n_0.eps}{algorithm_s_0.eps}{base_mergesort_0.eps}{double_aligned_mergesort_0.eps}{double_aligned_multi_mergsort_0.eps}{double_aligned_multi_mergsort_0.eps}{mergesorts}

>>>>

For his base algorithm, LaMarca takes Knuth's algorithm N \cite{Knuth98}.
LaMarca describes this algorithm as functioning as described above; lists of
length one are merged into lists of length two, then four, eight, and so on.
However, algorithm N does not work in this fashion. Instead, it works as a
\n{natural} merge. Rather than each list having a fixed length at a certain
point, lists are merged according to the initial ordering. If it happens that
there exists an ascending list on one side, this is exploited by the algorithm.
While this results in good performance, the merge is not perfectly regular, and
it is difficult or impossible to perform several of the optimizations LaMarca
describes on it. We attempted to contact him in order to discuss this, but
received no response. It is likely that the improvements were actually applied
to Knuth's algorithm S.

@@

\sixplot{algorithm_n_0.eps}{algorithm_s_0.eps}{base_mergesort_0.eps}{double_aligned_mergesort_0.eps}{double_aligned_multi_mergesort_0.eps}{double_aligned_multi_mergesort_0.eps}{mergesorts}

### 2005-07-31 docs/tech_report/software_predictor.tex
spell checking. just noticed the word mergsort, so need to fix it

\sixplot{base_quicksort_0.eps}{base_quicksort9_0.eps}{memory_tuned_quicksort_0.eps}{multi_quicksort_0.eps}{sequential_multi_quicksort_0.eps}{}{base, memory-tuned and multi quicksorts and base quicksort (pseudo-median-of-9)}
\sixplot{base_quicksort1_0.eps}{base_quicksort_0.eps}{base_quicksort5_0.eps}{base_quicksort7_0.eps}{base_quicksort9_0.eps}{}{base quicksort without a median, with median of 3, and with pseudo-medians of 5, 7 and 9}
\sixplot{algorithm_n_0.eps}{algorithm_n_1.eps}{algorithm_s_0.eps}{algorithm_s_1.eps}{}{}{Knuth's mergesorts}
\sixplot{base_mergesort_0.eps}{base_mergesort_1.eps}{base_mergesort_2.eps}{base_mergesort_3.eps}{base_mergesort_4.eps}{base_mergesort_5.eps}{base mergesort}
\sixplot{double_aligned_mergesort_0.eps}{double_aligned_mergesort_1.eps}{double_aligned_mergesort_2.eps}{double_aligned_mergesort_3.eps}{double_aligned_mergesort_4.eps}{double_aligned_mergesort_5.eps}{double-aligned tiled mergesort - part 1}
\sixplot{double_aligned_mergesort_6.eps}{double_aligned_mergesort_7.eps}{double_aligned_multi_mergsort_0.eps}{double_aligned_multi_mergsort_1.eps}{double_aligned_multi_mergsort_2.eps}{double_aligned_multi_mergsort_3.eps}{double-aligned tiled mergesort - part 2, and double multi mergesort - part 1}
\sixplot{double_aligned_multi_mergsort_4.eps}{double_aligned_multi_mergsort_5.eps}{double_aligned_multi_mergsort_6.eps}{double_aligned_multi_mergsort_7.eps}{double_aligned_multi_mergsort_8.eps}{double_aligned_multi_mergsort_9.eps}{double-aligned multi-mergesort - part 2}

>>>>

\sixplot{base_quicksort_0.eps}{base_quicksort9_0.eps}{memory_tuned_quicksort_0.eps}{multi_quicksort_0.eps}{sequential_multi_quicksort_0.eps}{}{base, memory-tuned and multi quicksorts and base quicksort (pseudo-median-of-9)}
\sixplot{base_quicksort1_0.eps}{base_quicksort_0.eps}{base_quicksort5_0.eps}{base_quicksort7_0.eps}{base_quicksort9_0.eps}{}{base quicksort without a median, with median of 3, and with pseudo-medians of 5, 7 and 9}
\sixplot{algorithm_n_0.eps}{algorithm_n_1.eps}{algorithm_s_0.eps}{algorithm_s_1.eps}{}{}{Knuth's mergesorts}
\sixplot{base_mergesort_0.eps}{base_mergesort_1.eps}{base_mergesort_2.eps}{base_mergesort_3.eps}{base_mergesort_4.eps}{base_mergesort_5.eps}{base mergesort}
\sixplot{double_aligned_mergesort_0.eps}{double_aligned_mergesort_1.eps}{double_aligned_mergesort_2.eps}{double_aligned_mergesort_3.eps}{double_aligned_mergesort_4.eps}{double_aligned_mergesort_5.eps}{double-aligned tiled mergesort - part 1}
\sixplot{double_aligned_mergesort_6.eps}{double_aligned_mergesort_7.eps}{double_aligned_multi_mergesort_0.eps}{double_aligned_multi_mergesort_1.eps}{double_aligned_multi_mergesort_2.eps}{double_aligned_multi_mergesort_3.eps}{double-aligned tiled mergesort - part 2, and double multi mergesort - part 1}
\sixplot{double_aligned_multi_mergesort_4.eps}{double_aligned_multi_mergesort_5.eps}{double_aligned_multi_mergesort_6.eps}{double_aligned_multi_mergesort_7.eps}{double_aligned_multi_mergesort_8.eps}{double_aligned_multi_mergesort_9.eps}{double-aligned multi-mergesort - part 2}
